---
title: "Simulations In-Class Project"
date: "Due October 13, 2017 at 11:59pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:



With this project we will simulate a famoues probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
```{r}
Gamblers_Ruin <- function(bet, bankroll, prob) {
  count <- 0
  while(bet >0) { 
  ProbTurn <- runif(1)
  if(ProbTurn <= prob) {
    bankroll <- bankroll + bet
    #now I need to raise the bet to a fixed fraction of the bankroll, given from the definition
    bet <- (bankroll + bet)*.1 # the raise is .1 since the first is 1/10 of the total bankroll (100/1000)
  } else {
    bankroll <- bankroll - bet
    bet <- bet
  }
  count <- count +1
  if(bankroll < bet) {break}
  }
  return(count)
}
```
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
```{r}
simulations <- vector()
for(i in 1:5000) {
  simulation <- Gamblers_Ruin(bet = 100, bankroll = 900, prob = 0.5)
  simulations <- c(simulations, simulation)
}
totalRuns <- length(simulations)
Under100 <- length(which(simulations<=100))
Prob <- Under100/totalRuns
Prob
# The Prob that you have "busted" by the time you placed your one hundredth bet is 0.1208, given that he raises his bet to a fixed fraction when he wins.
```
    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
```{r}
totalRuns <- length(simulations)
Under100 <- length(which(simulations<=500))
Prob <- Under100/totalRuns
Prob
# The probability that you have busted by the time you have placed your five hundredth bet is 0.4876.
```    
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
```{r}
under5000 <- simulations[which(simulations<=5000)]
mean(under5000)

# The mean time to go bust is 744.2552.
```    
    d. the mean and variance of your bankroll after 100 hands (including busts).
```{r}
Gamblers_Ruin2 <- function(bet, bankroll, prob) {
  count <- 0
  while(bet >0) { 
  ProbTurn <- runif(1)
  if(ProbTurn <= prob) {
    bankroll <- bankroll + bet
    #now I need to raise the bet to a fixed fraction of the bankroll, given from the definition
    bet <- (bankroll + bet)*.1 # the raise is .1 since the first is 1/10 of the total bankroll (100/1000)
  } else {
    bankroll <- bankroll - bet
    bet <- bet
  }
  count <- count +1
  if(count == 100) {break}
  if(bankroll < bet) {break}
  }
  return(bankroll)
}

bankrollAmounts <- vector()
for (i in 1:5000) {
  amount <- Gamblers_Ruin2(bet = 100, bankroll = 900, prob = 0.5)
  bankrollAmounts <- c(bankrollAmounts, amount)
}

mean(bankrollAmounts)
# The mean of the bankroll account after 100 hands is 902.5504.
var(bankrollAmounts)
# The variance of the bankroll account after 100 hands is 3194912.
```    
    e. the mean and variance of your bankroll after 500 hands (including busts).
```{r}
Gamblers_Ruin3 <- function(bet, bankroll, prob) {
  count <- 0
  while(bet >0) { 
  ProbTurn <- runif(1)
  if(ProbTurn <= prob) {
    bankroll <- bankroll + bet
    #now I need to raise the bet to a fixed fraction of the bankroll, given from the definition
    bet <- (bankroll + bet)*.1 # the raise is .1 since the first is 1/10 of the total bankroll (100/1000)
  } else {
    bankroll <- bankroll - bet
    bet <- bet
  }
  count <- count +1
  if(count == 500) {break}
  if(bankroll < bet) {break}
  }
  return(bankroll)
}

bankrollAmounts3 <- numeric()
for (i in 1:5000) {
  amount <- Gamblers_Ruin3(bet = 100, bankroll = 900, prob = 0.5)
  bankrollAmounts3 <- c(bankrollAmounts3, amount)
}

mean(bankrollAmounts3)
# The mean of the bankroll account after 500 hands is 750.2391.
var(bankrollAmounts3)
# The variance of the bankroll account after 500 hands is 130766877.
``` 
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

When the player goes bust, his bankroll will be less than the bet amount. When this is the case, the for loop breaks. I have this set up as an if statement.



2. Repeat the previous problem with betting on black in American roulette, where the probability of winning on any spin is 18/38 for an even payout.
```{r}
AmericanRoulette <- function(bankroll=900, bet=100, p=18/38) {
  count <- 0
  amount <- bankroll
  while(bankroll>0) {
    hand = amount + 2*bet*rbinom(1,1,p) - bet
    amount <- hand
    count <- count + 1
    if (amount == 0) {break}
  }
  return(count)
}

# Part (a): the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet.
test <- AmericanRoulette()
test

sims <- vector()
for (i in 1:5000) {
  hands <- AmericanRoulette()
  sims <- c(sims, hands)
}

Bust100 <- length(which(sims<=100))/length(sims)
# The probability of busting by the 100th hand is 55.46%

# Part (b): the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly.

Bust500 <- length(which(sims<=500))/length(sims)
# The probability of busting by the 500th hand is 93.38%

# Part (c): the mean time you go bust, given that you go bust within the first 5000 hands.

BustMean <- mean(sims[which(sims<=5000)])
# The mean time to bust is 167.8876.

# Part (d): the mean and variance of your bankroll after 100 hands (including busts).
AmericanRoulette2 <- function(bankroll=900, bet=100, p=18/38) {
  amount <- bankroll
  count <- 0
  while(bankroll>0) {
    hand = amount + 2*bet*rbinom(1,1,p) - bet
    amount <- hand
    count <- count +1
    if (amount == 0) {break}
    if (count ==100) {break}
  }
  return(amount)
}

sims3 <- vector()
for (i in 1:100) {
  hand <- AmericanRoulette2()
  sims3 <- c(sims3, hand)
}
mean(sims3)
var(sims3)

# The mean is 537 and the variance is 559,324.2


```

3. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
```{r}

Markov_Chains <- function(wager = 100, bankroll, p = 0.48) {
  prob <- p
  for (i in 1:100000) {
  ProbTurn <- runif(1) #determing the turn probability
  if(ProbTurn <= p) {
    bankroll <- bankroll + wager
    p = p + 0.01
  } else {
    bankroll <- bankroll - wager
    p <- prob
  }
  if(bankroll<wager) {break}
  }
  return(bankroll)
}

returns <- vector()
for(j in 1:100) {
  
  test <- Markov_Chains(wager = 100, 1000, p=0.48)
  returns <- c(returns, test)
}
mean(returns)
# This is not a fair game. The expected payout of simulations is 0 so the person making the bets will lose all of his bankroll in the longrun.
```
    
b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?

```{r}

returns2 <- vector()
for(i in 1:100) {
  test <- Markov_Chains(wager = 100, bankroll = 1000, p = 0.50)
  returns2 <- c(returns2, test)
}
mean(returns2)

# The starting probability of 0.5 is the fairest.
```

c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 
    
```{r}
Markov_Chains1 <- function(wager = 100, bankroll, p = 0.48) {
  prob <- p
  for (i in 1:100000) {
  ProbTurn <- runif(1) #determing the turn probability
  if(ProbTurn <= p) {
    bankroll <- bankroll + wager
    p = p + 0.025
  } else {
    bankroll <- bankroll - wager
    p <- prob
  }
  if(bankroll<wager) {break}
  }
  return(bankroll)
}

returns <- vector()
for(j in 1:100) {
  
  test <- Markov_Chains1(wager = 100, 1000, p=0.48)
  returns <- c(returns, test)
}
mean(returns)

```    
    


4. Creating a Bootstrap function. There is a particular concept called [bootstrapping]
(https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) where we can easily create 95% confidence intervals, even for complex estimators.

The steps of this process are:

  a. Draw a sample, with replacement, from your data which is the same length of your data.
  b. Calculate the statistic of interest on this boostrap sample (ie mean, variance, regression,...)
  c. Peform steps 1:2 at least 1000 times over until you have a vector of your statistics. 
  d. The lower bound of a 95% CI will be the 0.025 percentile
  e. The upper bound of a 95% CI will be the 0.975 percentile

Make a function called `boot_ci` which calculates the 95% confidence interval in this manner. 

5. For problems 3b and 3c, you calculated a mean value. Because you saved these final results in a vector, use the bootstrap to estimate the variance of the return in each case for your final answer. Once you have these results, which game has the smaller variance in returns?